Solving the independent set problem by generic programming tensor networks \thanks{\funding{...}}

### 2021-08-15 paper/paper.tex
update

\section{Introduction}
In this work, we introduce a tensor based framework to study the famous graph problem of finding independent sets.
Given an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set that for any $u,v \in I$, there is no edge connecting $u$ and $v$ in $G$.
The problem of finding the maximum independent set (MIS) size $\alpha(G) \equiv \max_{I}|I|$ belongs to the complexity class NP-complete~\cite{Hastad1996}, which is unlikely to be decided in polynomial time.
It is hard to even approximate this size in polynomial time within a factor $|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
The exhaustive search for a solution costs time $2^{|V|}$.
More efficient algorithms to compute the MIS size exactly includes the branching algorithm and dynamic programming.
Without changing the fact of exponential scaling in computing time, the branching algorithm gives a smaller base.
For example, in ~\cite{Xiao2017}, a sophisticated branching algorithm has a time complexity $1.1893^n n ^{O(1)}$.
The dynamic programming approach~\cite{Courcelle1990,Fomin2013} works better for graphs with small tree width $tw(G)$, it gives an algorithms of complexity $O(2^{tw(G)}tw(G)n)$.
People are interested in solving the independent set problem better not only because it is an NP-complete problem that directly related to other NP-complete problems like maximal cliques and vertex cover~\cite{Moore2011},
but also for its close relation with physical applications like hard spheres lattice gas model~\cite{Dyre2016}, and Rydberg hamiltonian~\cite{Pichler2018}.
However, in these applications, knowing the MIS size and one of the optimum solutions is not the only goal.
%A set is independent if and only if it is a clique in the graph's complement, so the two concepts are complementary.
%A set is independent if and only if its complement is a vertex cover. Therefore, the sum of the size of the largest independent set $\alpha (G)$ and the size of a minimum vertex cover $\beta (G)$ is equal to the number of vertices in the graph.
People often ask different questions about independent sets in order to understand the landscape of their models better.
These questions includes but not limited to, counting all independent sets, obtaining all independent sets of size $\alpha(G)$ and $\alpha(G)-1$,
counting the number of (maximal) independent sets of different sizes, and understanding the effect of a local gadget.
In this work, we attack this problem by mapping it to an generic tensor network.
It does not give a better time complexity comparing to dynamic programming, but is versatile enough to answer the above questions by engineering the tensor elements with minimum effort.

@@

Before contracting the tensor network and evaluating the independence polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A semiring is a ring without additive inverse, while a commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relations must hold for arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}

In the following, we show how to obtain the independence polynomial, the maximum independent set size and optimal configurations of a general graph $G$ by designing tensor element types as commutative semirings,
i.e. making the tensor network generic~\cite{Stepanov2014}.

\subsection{The polynomial approach}
A straight forward approach to evaluate the independence polynomial is treating the tensor elements as polynomials, and evaluate the polynomial directly.
Let us create a polynomial type, and represent a polynomial $a_0 + a_1 x + \ldots + a_k x^k$ as a vector $(a_0, a_1, \ldots, a_k) \in R^k$, e.g. $x$ is represented as $(0, 1)$.
We define the algebra between the polynomials $a$ of order $k_a$ and $b$ of order $k_b$ as
\begin{align}
    \begin{split}
    a \oplus b &= (a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_0 + b_0, a_1b_0 + a_0b_1, \ldots, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (),\\
    \mymathbb{1} &= (1).\label{eq:polynomial}
    \end{split}
\end{align}
By contracting the tensor network with polynomial type, the final result is the exact representation of the independence polynomial.
In the program, the multiplication can be evaluated efficiently with the convolution theorem~\cite{Schonhage1971}.
The only problem of this method is it suffers from a space overhead that proportional to the maximum independent set size because each polynomial requires a vector of such size to store the factors.
In the following subsections, we managed to solve this problem.

\subsection{The fitting and Fourier transformation approaches}
Let $m=\alpha(G)$ be the maximum independent set size and $X$ be a set of real numbers of cardinality $m+1$.
We compute the tensor network contraction for each $x_i \in X$ and obtain the following relations
\begin{align}
    \begin{split}
a_0 + a_1 x_1 + a_1 x_1^2 + \ldots + a_m x_1^m &= y_0\\
a_0 + a_1 x_2 + a_2 x_2^2 + \ldots + a_m x_2^m &= y_1\\
\ldots&\\
a_0 + a_1 x_m + a_2 x_m^2 + \ldots + a_m x_m^m& = y_m
    \end{split}
\end{align}
The polynomial fitting between $X$ and $Y = \{y_0, y_1, \ldots, y_m\}$ gives us the factors.
The polynomial fitting is essentially about solving the following linear equation
\begin{align}
\left(\begin{matrix}
1 & x_1 & x_1^2 & \ldots & x_1^m \\
1 & x_2 & x_2^2 & \ldots & x_2^m \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & x_m & x_m^2 & \ldots & x_m^m
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_m
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_m
\end{matrix}\right).\label{eq:lineareq}
\end{align}

\subsection{The finite field algebra approach}
It is possible but not trivial to compute the independence polynomial rigorously using integer number types only.
The hardness originates from the practical consideration of computing speed and precision.
Fixed width integer types are often too small to store the counting,
while big integer with varying width can be very slow and incompatible with graphic processing units (GPU) devices.
This problem can be solved by introducing finite field algebra $GF(p)$

\begin{align}
\begin{split}
    x ~\oplus~ y &= x+y\pmod p,\\
    x ~\odot~ y &= xy\pmod p,\\
    \mymathbb{0} &= 0,\\
    \mymathbb{1} &= 1.
\end{split}\label{eq:finitefield}
\end{align}

In a finite field algebra, we have the following observations
\begin{enumerate}
    \item One can use Gaussian elimination~\cite{Golub2013} to solve a linear equation \Eq{eq:lineareq}
    because it is a generic function that works for any elements with field algebra.
    The multiplicative inverse of a finite field algebra can be computed with the extended Euclidean algorithm.
    \item Given the remainders of a larger unkown integer $x$ over a set of co-prime integers $\{p_1, p_2, \ldots, p_n\}$,
    $x \pmod {p_1 \times p_2 \times \ldots \times p_n}$ can be computed using the Chinese remainder theorem.
    With this, one can infer big integers from small integers.
\end{enumerate}
With these observations, we developed Algorithm~\ref{alg:finitefield} to compute independence polynomial exactly without introducing space overheads.
In the algorithm, except the computation of Chinese remainder theorem, all computations are done with integers of fixed width $W$.

@@

\subsection{Maximal independence polynomial}
Some times people are interested in knowing maximal solutions to understand why their program is trapped in a local minimal.
Then they might want to compute the maximal independence polynomial.
Let us denote the neighbour of a vertex $v$ as $N(v)$ and $N[v] = N(v)\cup \{v\}$.
A maximal independent set $I_m$ is an independent sets that there does not exist a vertex $v$ that $N[v] \cap I_m = \emptyset$.
Let us modify the tensor network for computing independence polynomial by adding this restriction.
Instead of defining the restriction on vertices and edges, we define it on $N[v]$
\begin{align}
    T(x_v)_{s_1,s_2,\ldots,s_{|N(v)|},s_v} = \begin{cases}
        s_vx_v & s_1=s_2=\ldots=s_{|N(v)|}=0,\\
        1-s_v& otherwise.\\
    \end{cases}
\end{align}
As an example, for a vertex of degree 2, the resulting rank 3 tensor is
\begin{align}
    T(x_v)=\left(\begin{matrix}
    \left(\begin{matrix}
        0 &1 \\
        1 &1
    \end{matrix}\right)\\
    \left(\begin{matrix}
        x_v &0 \\
        0 &0
    \end{matrix}\right)
    \end{matrix}\right).
\end{align}

@@

\section{Enumerating configurations}
One may also want to obtain all solutions, it can be achieved replacing the factors $a_x$ with a set of bit strings $s_x$,
We design a new element type having the following algebra
\begin{align}
\begin{split}
    s \oplus t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau | \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes n}\}
\end{split}\label{eq:set}
\end{align}
\iffalse
\begin{align}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes n}\}\infty^{0},
\end{split}
\end{align}
\fi
where $\lor^\circ$ is the Hadamard logic or operation over two bit strings, which means joining of two local configurations.
The variable $x$ in the vertex tensor is initialized to $x_i = \{\boldsymbol{e}_{i}\}$,
where $\boldsymbol{e}_i$ is a one hot vector of size $|G|$.
One can easily check this algebra is a commutative semiring.
When we use the above algebra as factors of independence polynomials, the resulting algebra is also a commutative semiring.
With this new element type, the tensor network contraction will give all solutions rather than just a number for counting.
By slightly modifying the above algebra, it can also be used to obtain just a single configuration to save the computational effort.
%We leave this as an exercise for readers.

@@

\subsection{Bounding the enumeration space}
When one uses the set algebra in \Eq{eq:set} to represent the factors in \Eq{eq:countingtropical} for enumerating all optimum configurations, he will find the program stores more than necessary intermediate configurations and cause significant overheads in space.
To speed up the computation, we use $\alpha(G)$ to bound the searching space.
We first compute the value of $\alpha(G)$ with tropical numbers and cache all intermediate tensors.
Then we compute a boolean masks for each cached tensor, where we use a boolean true to represent a tensor element having contribution to the maximum independent set (i.e. with a non-zero gradient) and boolean false otherwise.
Finally, we perform masked matrix multiplication using the new element type with the above algebra for obtaining all configurations.
Notice that these masks are in fact tensor elements with non-zero gradients with respect to MIS size, we compute these masks by back propagating gradients.
To derive the backward rule, we consider a tropical matrix multiplication $C = A B$, we have the following inequality
\begin{align}
    A_{ij} \odot B_{jk} &\leq C_{ik}.
\end{align}
Moving $B_{ik}$ to the right hand side, we have
\begin{align}
    A_{ij} &\leq (\oplus_{k} (C_{ik}^{-1} \odot B_{jk}))^{-1}
\end{align}
where the tropical multiplicative inverse is defined as the additive inverse of the regular algebra. The equality holds if and only if element $A_{ij}$ contributions to $C$ (i.e. has non-zero gradient).
Let the mask for $C$ being $\overline C$, the backward rule for ``gradient'' masks reads
\begin{align}
\overline{A}_{ij} = \delta(A_{ij}, ((C^{\circ-1} \circ \overline C )B^T)_{ij}^{\circ -1}),
\end{align}
where ${}^{\circ -1}$ is the Hadamard inverse, $\circ$ is the Hadamard product, boolean false is treated as tropical zero and boolean true is treated as tropical one.
This rule defined on matrix multiplication can be easily generalized to tensor contraction by replacing the matrix multiplication between $C^{\circ-1} \circ \overline C$ and $B^T$ by a tensor contraction.~\cite{} \blue{maybe add an appendix?}  % my blog

@@

>>>>

\section{Introduction}
\blue{efficient counting maximal independent sets~\cite{Manne2013}}
In this work, we introduce a tensor based framework to study the famous graph problem of finding independent sets.
Given an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set that for any $u,v \in I$, there is no edge connecting $u$ and $v$ in $G$.
The problem of finding the maximum independent set (MIS) size $\alpha(G) \equiv \max_{I}|I|$ belongs to the complexity class NP-complete~\cite{Hastad1996}, which is unlikely to be decided in polynomial time.
It is hard to even approximate this size in polynomial time within a factor $|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
The exhaustive search for a solution costs time $2^{|V|}$.
More efficient algorithms to compute the MIS size exactly includes the branching algorithm and dynamic programming.
Without changing the fact of exponential scaling in computing time, the branching algorithm gives a smaller base.
For example, in ~\cite{Xiao2017}, a sophisticated branching algorithm has a time complexity $1.1893^n n ^{O(1)}$.
The dynamic programming approach~\cite{Courcelle1990,Fomin2013} works better for graphs with small tree width $tw(G)$, it gives an algorithms of complexity $O(2^{tw(G)}tw(G)n)$.
People are interested in solving the independent set problem better not only because it is an NP-complete problem that directly related to other NP-complete problems like maximal cliques and vertex cover~\cite{Moore2011},
but also for its close relation with physical applications like hard spheres lattice gas model~\cite{Dyre2016}, and Rydberg hamiltonian~\cite{Pichler2018}.
However, in these applications, knowing the MIS size and one of the optimum solutions is not the only goal.
%A set is independent if and only if it is a clique in the graph's complement, so the two concepts are complementary.
%A set is independent if and only if its complement is a vertex cover. Therefore, the sum of the size of the largest independent set $\alpha (G)$ and the size of a minimum vertex cover $\beta (G)$ is equal to the number of vertices in the graph.
People often ask different questions about independent sets in order to understand the landscape of their models better.
These questions includes but not limited to, counting all independent sets, obtaining all independent sets of size $\alpha(G)$ and $\alpha(G)-1$,
counting the number of (maximal) independent sets of different sizes, and understanding the effect of a local gadget.
In this work, we attack this problem by mapping it to an generic tensor network.
It does not give a better time complexity comparing to dynamic programming, but is versatile enough to answer the above questions by engineering the tensor elements with minimum effort.

@@

Before contracting the tensor network and evaluating the independence polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A semiring is a ring without additive inverse, while a commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relations must hold for arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
The property of being commutative is required here because we want the contraction result independent of the contraction order.
In the following, we show how to obtain the independence polynomial, the maximum independent set size and optimal configurations of a general graph $G$ by designing tensor element types as commutative semirings,
i.e. making the tensor network generic~\cite{Stepanov2014}.
A straight forward approach to evaluate the independence polynomial is treating the tensor elements as polynomials, and evaluate the polynomial directly.
Let us create a polynomial type, and represent a polynomial $a_0 + a_1 x + \ldots + a_k x^k$ as a vector $(a_0, a_1, \ldots, a_k) \in R^k$, e.g. $x$ is represented as $(0, 1)$.
We define the algebra between the polynomials $a$ of order $k_a$ and $b$ of order $k_b$ as
\begin{align}
    \begin{split}
    a \oplus b &= (a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_0 + b_0, a_1b_0 + a_0b_1, \ldots, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (),\\
    \mymathbb{1} &= (1).\label{eq:polynomial}
    \end{split}
\end{align}
By contracting the tensor network with polynomial type, the final result is the exact representation of the independence polynomial.
In the program, the multiplication can be evaluated efficiently with the convolution theorem~\cite{Schonhage1971}.
However, this approach suffers from a space overhead that proportional to the maximum independent set size because each polynomial requires a vector of such size to store the factors.
In \App{app:fft}, we provide a fitting based approach to compute the independence polynomial.
One just set $x$ to $\alpha(G)+1$ random values, compute the result and use the polynomial fitting to get the factors.
In this way, we do not have linear overheads in space, however, 
due to fact that countings of different MIS sizes can be different in many orders, 
the round off error dominates the high MIS size region that we are most interested about if we use floating point numbers in computation, 
meanwhile the number easily overflows if we use fixed width integer types.
The big integer type is not an option because big integers with varying width can be very slow and incompatible with graphic processing units (GPU) devices.
Then we want to resort to integer numbers, however, fixed width integer types are often too small to store the counting,
This problem can be solved by introducing finite field algebra $GF(p)$
\begin{align}
\begin{split}
    x ~\oplus~ y &= x+y\pmod p,\\
    x ~\odot~ y &= xy\pmod p,\\
    \mymathbb{0} &= 0,\\
    \mymathbb{1} &= 1.
\end{split}\label{eq:finitefield}
\end{align}
In a finite field algebra, we have the following observations
\begin{enumerate}
    \item One can use Gaussian elimination~\cite{Golub2013} to solve a linear equation \Eq{eq:lineareq}
    because it is a generic function that works for any elements with field algebra.
    The multiplicative inverse of a finite field algebra can be computed with the extended Euclidean algorithm.
    \item Given the remainders of a larger unkown integer $x$ over a set of co-prime integers $\{p_1, p_2, \ldots, p_n\}$,
    $x \pmod {p_1 \times p_2 \times \ldots \times p_n}$ can be computed using the Chinese remainder theorem.
    With this, one can infer big integers from small integers.
\end{enumerate}
With these observations, we developed Algorithm~\ref{alg:finitefield} to compute independence polynomial exactly without introducing space overheads.
In the algorithm, except the computation of Chinese remainder theorem, all computations are done with integers of fixed width $W$.

@@

\subsection{Maximal independence polynomial}
Some times people are interested in knowing maximal solutions to understand why their programs are trapped in a local minimal.
Then they might want to compute the maximal independence polynomial.
Let us denote the neighbour of a vertex $v$ as $N(v)$ and $N[v] = N(v)\cup \{v\}$.
A maximal independent set $I_m$ is an independent sets that there does not exist a vertex $v$ that $N[v] \cap I_m = \emptyset$.
Let us modify the tensor network for computing independence polynomial by adding this restriction.
Instead of defining the restriction on vertices and edges, we define it on $N[v]$
\begin{align}
    T(x_v)_{s_1,s_2,\ldots,s_{|N(v)|},s_v} = \begin{cases}
        s_vx_v & s_1=s_2=\ldots=s_{|N(v)|}=0,\\
        1-s_v& otherwise.\\
    \end{cases}
\end{align}
As an example, for a vertex of degree 2, the resulting rank 3 tensor is
\begin{align}
    T(x_v)=\left(\begin{matrix}
    \left(\begin{matrix}
        0 &1 \\
        1 &1
    \end{matrix}\right)\\
    \left(\begin{matrix}
        x_v &0 \\
        0 &0
    \end{matrix}\right)
    \end{matrix}\right).
\end{align}

@@

\section{Enumerating configurations}
One may also want to obtain all solutions, it can be achieved replacing the factors with a set of solutions,
We design a new element type having the following algebra
\begin{align}
\begin{split}
    s \oplus t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau | \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes n}\}
\end{split}\label{eq:set}
\end{align}
where $\lor^\circ$ is the Hadamard logic or operation over two bit strings, which means joining of two local configurations.
The variable $x$ in the vertex tensor is initialized to $x_i = \{\boldsymbol{e}_{i}\}$,
where $\boldsymbol{e}_i$ is a one hot vector of size $|G|$.
As an example, if we want to enumerate all maximum independent sets, we can use the above set as the factors in \Eq{eq:countingtropical}.
We get the following new algebra.
\begin{align}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes n}\}\infty^{0},
\end{split}
\end{align}
One can easily check if the factor algebra is a commutative semiring,
when we use the above algebra as factors of independence polynomials, the resulting algebra is also a commutative semiring.
If one is only interested in obtaining a single configuration, one can also just keep a single configuration to save the computational effort.
We arrive at a new algebra defined on bit strings.
%We leave this as an exercise for readers.

@@

\subsection{Bounding the enumeration space}
When one uses the set algebra in \Eq{eq:set} to represent the factors in \Eq{eq:countingtropical} for enumerating all optimum configurations, he will find the program stores more than necessary intermediate configurations and cause significant overheads in space.
To speed up the computation, we use $\alpha(G)$ to bound the searching space.
We first compute the value of $\alpha(G)$ with tropical numbers and cache all intermediate tensors.
Then we compute a boolean masks for each cached tensor, where we use a boolean true to represent a tensor element having contribution to the maximum independent set (i.e. with a non-zero gradient) and boolean false otherwise.
Finally, we perform masked matrix multiplication using the new element type with the above algebra for obtaining all configurations.
Notice that these masks are in fact tensor elements with non-zero gradients with respect to MIS size, we compute these masks by back propagating gradients.
To derive the backward rule for tensor contraction,
we first reduce the problem to finding the backward rule of a tropical matrix multiplication $C = A B$,
where we have the following inequality
\begin{align}
    A_{ij} \odot B_{jk} &\leq C_{ik}.
\end{align}
Here is $\leq$ on tropical numbers are same as regular algebra.
The tropical multiplication $\odot$ is the same as the regular $+$, then
one can move $B_{ik}$ to the right hand side and get
\begin{align}
    A_{ij} &\leq C_{ik} \odot B_{jk}^{\circ -1}
\end{align}
where the tropical multiplicative inverse is defined as the additive inverse of the regular algebra.
The inequality still holds if we take the minimum over $k$
\begin{align}
    \begin{split}
    A_{ij} &\leq \min_{k}(C_{ik} \odot B_{jk}^{\circ -1}) = (\oplus_{k} (C_{ik}^{-1} \odot B_{jk}))^{\circ -1}.
    \end{split}
\end{align}
On the right hand side, we transformed the operation into a tropical matrix multiplication so that we can utilize the fast tropical BLAS routines.
The equality holds if and only if element $A_{ij}$ has contribution to $C$ (i.e. has non-zero gradient).
Let the gradient mask for $C$ being $\overline C$, the backward rule for gradient masks reads
\begin{align}
\overline{A}_{ij} = \delta(A_{ij}, ((C^{\circ-1} \circ \overline C )B^T)_{ij}^{\circ -1}),
\end{align}
where ${}^{\circ -1}$ is the Hadamard inverse, $\circ$ is the Hadamard product, boolean false is treated as tropical zero and boolean true is treated as tropical one.
This rule defined on matrix multiplication can be easily generalized to tensor contraction by replacing the matrix multiplication between $C^{\circ-1} \circ \overline C$ and $B^T$ by a tensor contraction.~\cite{} \blue{maybe add an appendix?}  % my blog

@@

\section{The fitting and Fourier transformation approaches to computing independence polynomial}\label{app:fft}
Let $m=\alpha(G)$ be the maximum independent set size and $X$ be a set of real numbers of cardinality $m+1$.
We compute the tensor network contraction for each $x_i \in X$ and obtain the following relations
\begin{align}
    \begin{split}
a_0 + a_1 x_1 + a_1 x_1^2 + \ldots + a_m x_1^m &= y_0\\
a_0 + a_1 x_2 + a_2 x_2^2 + \ldots + a_m x_2^m &= y_1\\
\ldots&\\
a_0 + a_1 x_m + a_2 x_m^2 + \ldots + a_m x_m^m& = y_m
    \end{split}
\end{align}
The polynomial fitting between $X$ and $Y = \{y_0, y_1, \ldots, y_m\}$ gives us the factors.
The polynomial fitting is essentially about solving the following linear equation
\begin{align}
\left(\begin{matrix}
1 & x_1 & x_1^2 & \ldots & x_1^m \\
1 & x_2 & x_2^2 & \ldots & x_2^m \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & x_m & x_m^2 & \ldots & x_m^m
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_m
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_m
\end{matrix}\right).\label{eq:lineareq}
\end{align}
